---
title: "Fitting diagnostics: convergence, residuals, and uncertainty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting diagnostics: convergence, residuals, and uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
```

```{r packages, message=FALSE}
library(tidysoilwater)
library(dplyr)
library(tibble)
library(ggplot2)
```

## Overview

`fit_swrc()` returns a convergence flag and standard errors alongside the
fitted parameters. This vignette shows how to:

1. Check and handle non-converged fits
2. Plot residuals to detect systematic model misfit
3. Interpret parameter standard errors and compute confidence intervals
4. Understand starting-value sensitivity, particularly for clay soils

---

## 1. Detecting non-converged fits

The `convergence` column is `TRUE` when `nls()` converged and `FALSE`
(with `NA` parameters) when it did not.

```{r setup-data}
set.seed(2947)

# Reference parameters for three soil types
soils <- tribble(
  ~soil,          ~theta_r, ~theta_s, ~alpha,   ~n,
  "sandy loam",      0.047,    0.430,  0.145, 2.68,
  "loam",            0.078,    0.430,  0.036, 1.56,
  "silty clay loam", 0.089,    0.430,  0.010, 1.23   # difficult for NLS
)

heads   <- c(0, 10, 33, 100, 330, 1000, 3300, 5000, 15000)
sigma   <- 0.005

# Generate synthetic observations
obs <- soils |>
  tidyr::crossing(h = heads) |>
  mutate(
    m     = 1 - 1 / n,
    theta_true = theta_r + (theta_s - theta_r) /
                   (1 + (alpha * h)^n)^m,
    theta = pmax(0.01, theta_true + rnorm(n(), 0, sigma))
  ) |>
  select(soil, h, theta)
```

```{r fit-all}
fits <- obs |>
  group_by(soil) |>
  fit_swrc(theta_col = theta, h_col = h)

fits |>
  select(soil, theta_r, theta_s, alpha, n, convergence)
```

```{r convergence-check}
n_failed <- sum(!fits$convergence)
cat(sprintf("%d / %d fits converged.\n", sum(fits$convergence), nrow(fits)))

if (n_failed > 0) {
  cat("\nNon-converged soils:\n")
  fits |>
    filter(!convergence) |>
    pull(soil) |>
    cat(sep = "\n")
}
```

**What to do when a fit fails:**

* Try different starting values (see Section 4)
* Add more observations near the dry end (high |h|) where the curve flattens
* For very fine-textured soils (n < 1.2), consider fixing Оё_s from bulk
  density measurements and fitting only Оё_r, Оұ, n

---

## 2. Residual analysis

A good fit should show residuals scattered randomly around zero. Systematic
patterns suggest model misfit or bad observations.

```{r residuals}
# Reconstruct fitted curves and compute residuals
predicted <- fits |>
  select(soil, theta_r, theta_s, alpha, n) |>
  filter(!is.na(theta_r)) |>
  tidyr::crossing(h = heads) |>
  mutate(m = 1 - 1 / n) |>
  swrc_van_genuchten(theta_r = theta_r, theta_s = theta_s,
                     alpha = alpha, n = n, h = h) |>
  rename(theta_fitted = .theta)

resid_df <- obs |>
  left_join(predicted |> select(soil, h, theta_fitted),
            by = c("soil", "h")) |>
  filter(!is.na(theta_fitted)) |>
  mutate(residual = theta - theta_fitted)
```

```{r residual-plot}
ggplot(resid_df, aes(x = h, y = residual, colour = soil)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(size = 2) +
  scale_x_log10(labels = scales::label_comma()) +
  facet_wrap(~soil) +
  labs(
    x     = "Matric potential |h| (cm)",
    y     = expression(theta[obs] - theta[fit]),
    title = "Residuals by soil type"
  ) +
  theme(legend.position = "none")
```

```{r rmse}
resid_df |>
  group_by(soil) |>
  summarise(
    RMSE  = sqrt(mean(residual^2)),
    MAE   = mean(abs(residual)),
    max_e = max(abs(residual))
  )
```

---

## 3. Parameter standard errors and confidence intervals

`fit_swrc()` returns standard errors for all four parameters. For approximate
95% CIs use Вұ 2 SE; for exact intervals use `confint()` on the `nls` object
directly (see note below).

```{r se-table}
fits |>
  filter(convergence) |>
  select(soil, theta_r, std_error_theta_r, alpha, std_error_alpha, n, std_error_n)
```

```{r ci-approximate}
# 95 % CI: estimate Вұ 1.96 Г— SE
fits |>
  filter(convergence) |>
  mutate(
    n_lo = n - 1.96 * std_error_n,
    n_hi = n + 1.96 * std_error_n
  ) |>
  select(soil, n, n_lo, n_hi, std_error_n)
```

**Interpretation:**

* Narrow SE relative to estimate вҶ’ parameter is well-constrained by the data
* SE larger than ~25% of estimate вҶ’ consider adding observations at extremes
  of the retention curve (very wet or very dry)
* The `n` parameter typically has the smallest relative SE when the data span
  a wide pressure range (several log-cycles)

---

## 4. Starting value sensitivity

Default starting values (`Оё_r = 0.9 Г— min(Оё)`, `Оё_s = 1.05 Г— max(Оё)`,
`Оұ = 0.1`, `n = 2.0`) work well for most mineral soils with n вүҘ 1.4.
Clay soils (n вүҲ 1.1вҖ“1.3) can require tighter, soil-specific starts.

```{r sensitivity-demo}
# Difficult clay loam: n вүҲ 1.23
clay_obs <- obs |> filter(soil == "silty clay loam")

# Strategy: grid search over (alpha, n) starting points
start_grid <- expand.grid(
  alpha_start = c(0.005, 0.01, 0.02, 0.05),
  n_start     = c(1.1, 1.3, 1.5, 2.0)
)

fit_with_start <- function(a0, n0, obs_df) {
  theta_vec <- obs_df$theta
  h_vec     <- abs(obs_df$h)

  tryCatch({
    fit <- nls(
      theta ~ theta_r + (theta_s - theta_r) /
                (1 + (alpha * h_vec)^n)^(1 - 1/n),
      data  = list(theta = theta_vec, h_vec = h_vec),
      start = list(theta_r = min(theta_vec) * 0.9,
                   theta_s = max(theta_vec) * 1.05,
                   alpha   = a0,
                   n       = n0),
      control = nls.control(maxiter = 200)
    )
    broom::tidy(fit) |>
      tidyr::pivot_wider(names_from = term, values_from = estimate) |>
      mutate(alpha_start = a0, n_start = n0, converged = TRUE)
  },
  error = function(e)
    tibble(alpha_start = a0, n_start = n0, converged = FALSE)
  )
}

start_results <- purrr::map2_dfr(
  start_grid$alpha_start,
  start_grid$n_start,
  fit_with_start,
  obs_df = clay_obs
)

start_results |>
  select(alpha_start, n_start, converged, alpha, n) |>
  arrange(desc(converged), n_start)
```

### Key recommendations for clay soils

1. **Оұ starting point**: use 0.005вҖ“0.02 (reflecting fine-textured pores)
2. **n starting point**: stay in the expected range 1.1вҖ“1.3 rather than
   defaulting to 2.0
3. **Data requirements**: at minimum 8вҖ“10 observations spanning pF 0вҖ“4.2;
   include at least two points between pF 2вҖ“3 where clay SWRCs bend
4. **Fixed parameters**: if K_sat is measured, pass it as Оё_s constraint;
   reduces the fitting problem to three parameters

---

## 5. Visualising fitted curves alongside observations

```{r overlay-plot}
# Only converged soils
fit_curves <- fits |>
  filter(convergence) |>
  select(soil, theta_r, theta_s, alpha, n) |>
  tidyr::crossing(h = 10^seq(0, 5, by = 0.1)) |>
  mutate(m = 1 - 1 / n) |>
  swrc_van_genuchten(theta_r = theta_r, theta_s = theta_s,
                     alpha = alpha, n = n, h = h)

ggplot() +
  geom_line(data  = fit_curves,
            aes(x = h, y = .theta, colour = soil),
            linewidth = 0.9) +
  geom_point(data = obs |> filter(soil != "silty clay loam"),
             aes(x = h, y = theta, colour = soil),
             size = 2.5) +
  scale_x_log10(labels = scales::label_comma()) +
  labs(
    x      = "Matric potential |h| (cm)",
    y      = expression(theta~(m^3/m^3)),
    colour = "Soil",
    title  = "Observed (points) vs fitted VG curves"
  )
```

---

## Summary

| Diagnostic | What to check |
|------------|---------------|
| `convergence` column | `FALSE` вҶ’ adjust starting values or data range |
| Residual plot | Systematic patterns вҶ’ model misfit; outliers вҶ’ bad observations |
| `std_error_*` columns | SE / estimate > 0.25 вҶ’ more data needed |
| Starting value grid | Clay soils: try Оұ вҲҲ [0.005, 0.02], n вҲҲ [1.1, 1.3] |
