---
title: "Unsaturated hydraulic conductivity with hydraulic_conductivity()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Unsaturated hydraulic conductivity with hydraulic_conductivity()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
```

```{r packages, message=FALSE}
library(tidysoilwater)
library(dplyr)
library(tibble)
library(ggplot2)
```

## Overview

`hydraulic_conductivity()` computes unsaturated hydraulic conductivity K(h)
using the Mualem (1976) – Van Genuchten (1980) model. The effective saturation
S_e is first computed from the Van Genuchten retention parameters:

$$S_e(h) = \frac{1}{\left(1 + (\alpha |h|)^n\right)^m}$$

Then K(h) is:

$$K(h) = K_s \cdot S_e^{0.5} \cdot \left(1 - \left(1 - S_e^{1/m}\right)^m\right)^2$$

where m = 1 − 1/n (the Mualem constraint). Note that `m` must be supplied
explicitly to `hydraulic_conductivity()` — it is not inferred from `n`.

---

## 1. Parameters

| Parameter | Symbol | Units | Notes |
|-----------|--------|-------|-------|
| `ks`    | K_s | cm/day (or any rate) | Saturated hydraulic conductivity |
| `alpha` | α   | 1/cm | Same unit system as `h` |
| `n`     | n   | –    | Must be > 1 |
| `m`     | m   | –    | Typically `1 - 1/n`; must satisfy 0 < m < 1 |
| `h`     | h   | cm   | Absolute value used internally |

---

## 2. Single-soil K(h) curve

```{r single-kh}
# Loam soil
loam <- tibble(h = c(0, 10, 33, 100, 330, 1000, 3300, 10000, 15000))

loam_kh <- hydraulic_conductivity(loam,
  ks    = 10.4,          # cm/day (Carsel & Parrish 1988)
  alpha = 0.036,
  n     = 1.56,
  m     = 1 - 1 / 1.56,
  h     = h
)

loam_kh
```

```{r single-kh-plot}
ggplot(loam_kh, aes(x = h, y = .K)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_log10(labels = scales::label_comma()) +
  scale_y_log10(labels = scales::label_scientific()) +
  labs(
    x     = "Matric potential |h| (cm)",
    y     = "K(h) (cm/day)",
    title = "Unsaturated hydraulic conductivity — loam"
  )
```

---

## 3. Comparing texture classes

Standard Carsel & Parrish (1988) parameters. K_sat values span several
orders of magnitude from clay to sand.

```{r texture-params}
textures <- tribble(
  ~texture,        ~ks,   ~alpha,   ~n,
  "sand",          712.8,  0.145, 2.68,
  "sandy loam",    106.1,  0.075, 1.89,
  "loam",           10.4,  0.036, 1.56,
  "clay loam",       2.2,  0.019, 1.31,
  "clay",            4.8,  0.015, 1.09
)
```

```{r texture-kh-curves}
heads <- 10^seq(0, 5, by = 0.1)

kh_curves <- textures |>
  mutate(m = 1 - 1 / n) |>
  tidyr::crossing(h = heads) |>
  hydraulic_conductivity(ks = ks, alpha = alpha, n = n, m = m, h = h)

ggplot(kh_curves, aes(x = h, y = .K, colour = texture)) +
  geom_line(linewidth = 0.9) +
  scale_x_log10(labels = scales::label_comma()) +
  scale_y_log10(labels = scales::label_scientific()) +
  labs(
    x      = "Matric potential |h| (cm)",
    y      = "K(h) (cm/day)",
    colour = "Texture",
    title  = "Unsaturated hydraulic conductivity by texture class"
  )
```

---

## 4. K(h) in context: paired SWRC and K(h) plots

K(h) and the SWRC are derived from the same parameter set. Viewing them
together reveals the relationship between water storage and flow capacity.

```{r paired-plots, fig.height=8}
loam_combined <- tibble(h = heads) |>
  mutate(m = 1 - 1 / 1.56) |>
  swrc_van_genuchten(theta_r = 0.078, theta_s = 0.430,
                     alpha = 0.036, n = 1.56, h = h) |>
  hydraulic_conductivity(ks = 10.4, alpha = 0.036,
                          n = 1.56, m = m, h = h)

p_theta <- ggplot(loam_combined, aes(x = h, y = .theta)) +
  geom_line(linewidth = 1, colour = "steelblue") +
  scale_x_log10(labels = scales::label_comma()) +
  labs(x = NULL, y = expression(theta~(m^3/m^3)),
       title = "Loam: SWRC (top) and K(h) (bottom)")

p_k <- ggplot(loam_combined, aes(x = h, y = .K)) +
  geom_line(linewidth = 1, colour = "darkorange") +
  scale_x_log10(labels = scales::label_comma()) +
  scale_y_log10(labels = scales::label_scientific()) +
  labs(x = "|h| (cm)", y = "K(h) (cm/day)")

# Stack with patchwork if available, otherwise print separately
if (requireNamespace("patchwork", quietly = TRUE)) {
  library(patchwork)
  p_theta / p_k
} else {
  print(p_theta)
  print(p_k)
}
```

---

## 5. K_sat comparison table

```{r ksat-table}
textures |>
  mutate(m = 1 - 1 / n) |>
  arrange(desc(ks)) |>
  select(texture, ks, alpha, n) |>
  rename(`K_sat (cm/day)` = ks,
         `α (1/cm)` = alpha)
```

Note that K_sat for sand is about 150× higher than for loam, and ~650× higher
than for clay loam. Even a small change in texture classification can
represent a large hydraulic contrast.

---

## 6. Tidy evaluation: columns or scalars

Like `swrc_van_genuchten()`, all parameters accept bare column names or
scalars.

```{r tidyeval}
# Vary K_sat and alpha across rows; scalar n and m
tibble(
  site = c("ridge", "midslope", "valley"),
  ks   = c(150, 30, 5),
  alpha = c(0.10, 0.04, 0.015)
) |>
  tidyr::crossing(h = c(10, 100, 1000)) |>
  hydraulic_conductivity(ks = ks, alpha = alpha,
                          n = 1.6, m = 1 - 1/1.6, h = h) |>
  select(site, h, alpha, ks, .K)
```

---

## 7. Performance

```{r perf}
df_1m <- tibble(h = rep(c(33, 330, 15000), length.out = 1e6))
elapsed <- system.time(
  hydraulic_conductivity(df_1m, ks = 10, alpha = 0.036,
                          n = 1.56, m = 1 - 1/1.56, h = h)
)["elapsed"]
cat(sprintf("1 M rows in %.0f ms\n", elapsed * 1000))
```

---

## See also

* `vignette("soil-water-retention")` — the SWRC function that shares these parameters
* `vignette("fitting-diagnostics")` — fit parameters from observed (h, θ) data
* `vignette("raster-workflow")` — map K(h) across a spatial grid
