---
title: "Multi-pedon workflow: fitting retention curves at scale"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-pedon workflow: fitting retention curves at scale}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
have_sf     <- requireNamespace("sf",     quietly = TRUE)
have_scales <- requireNamespace("scales", quietly = TRUE)
have_tidyr  <- requireNamespace("tidyr",  quietly = TRUE)
```

## Overview

In landscape-scale soil hydrology studies, Van Genuchten parameters must be
fitted independently at each sampling location (pedon).  With hundreds or
thousands of pedons, sequential NLS fitting becomes a throughput bottleneck.
This vignette demonstrates:

1. Constructing a realistic multi-pedon dataset from synthetic field
   observations.
2. Running `fit_swrc()` sequentially and in parallel with
   `workers = parallel::detectCores(logical = FALSE)`.
3. Mapping the fitted saturated water content (θ_s) and shape parameter (*n*)
   back to pedon locations.

---

## Setup

```{r packages}
library(tidysoilwater)
library(dplyr)
library(ggplot2)
```

---

## 1  Generate a synthetic pedon network

We simulate a grid of 200 pedons spread across a hypothetical study area.
Each pedon is assigned one of four dominant soil textures (loam, sandy loam,
clay loam, or silt loam) based on its position, mimicking the kind of
spatial structure you would expect from a soil survey.

```{r pedon-grid}
set.seed(2047)

# Van Genuchten parameters from Van Genuchten (1980, Table 1)
# Van Genuchten parameters for four textures with n > 1.5 — these are
# well-conditioned for NLS fitting.  Soils with n close to 1 (e.g. clay,
# silty clay loam) have flat retention curves that make NLS poorly
# conditioned; a real workflow would require more robust starting values
# or a Bayesian approach for those classes.
vg_reference <- tribble(
  ~texture,      ~theta_r, ~theta_s, ~alpha,  ~n,
  "sand",        0.045,    0.430,    0.145,   2.68,
  "loamy sand",  0.057,    0.410,    0.124,   2.28,
  "sandy loam",  0.065,    0.410,    0.075,   1.89,
  "loam",        0.078,    0.430,    0.036,   1.56
)

# 200 pedons on a 20 × 10 spatial grid (arbitrary units)
pedons <- expand.grid(x = seq(1, 20), y = seq(1, 10)) |>
  as_tibble() |>
  mutate(
    pedon_id = row_number(),
    # Texture follows a simple catena: coarser soils in the west,
    # finer (loam) in the east — a typical hillslope pattern
    texture = case_when(
      x <= 5  ~ "sand",
      x <= 10 ~ "loamy sand",
      x <= 15 ~ "sandy loam",
      TRUE    ~ "loam"
    )
  ) |>
  left_join(vg_reference, by = "texture")

glimpse(pedons)
```

---

## 2  Simulate field observations

Laboratory or field tensiometer measurements typically cover 8–12 pressure
heads from saturation (h = 0) to permanent wilting point (h = 15 000 cm).
We generate synthetic θ(h) observations by evaluating the VG model at each
pedon's true parameters, then adding realistic measurement noise (σ = 0.005
m³/m³).

```{r simulate-obs}
h_obs <- c(0, 1, 5, 10, 50, 100, 300, 500, 1000, 3000, 5000, 15000)

# Each row of pedons becomes a set of (h, theta) observations
obs_data <- pedons |>
  rowwise() |>
  mutate(
    # Slight spatial heterogeneity: jitter parameters within ±10 %
    alpha_loc   = alpha   * runif(1, 0.90, 1.10),
    n_loc       = max(1.05, n * runif(1, 0.90, 1.10)),
    theta_r_loc = theta_r * runif(1, 0.90, 1.10),
    theta_s_loc = theta_s * runif(1, 0.90, 1.10),
    obs = list({
      theta_true <- swrc_van_genuchten(
        tibble(h = h_obs),
        theta_r = theta_r_loc, theta_s = theta_s_loc,
        alpha   = alpha_loc,   n       = n_loc,
        h       = h
      )$.theta
      tibble(
        h     = h_obs,
        theta = pmax(0, theta_true + rnorm(length(h_obs), sd = 0.005))
      )
    })
  ) |>
  tidyr::unnest(obs) |>
  ungroup()

cat(sprintf("%d pedons × %d pressure heads = %d observations total\n",
            n_distinct(obs_data$pedon_id),
            length(h_obs),
            nrow(obs_data)))
```

Spot-check two pedons from contrasting texture zones:

```{r spot-check, fig.alt="Retention curves for two pedons with contrasting soil textures"}
filter(obs_data, pedon_id %in% c(3, 178)) |>
  mutate(label = paste0("Pedon ", pedon_id, " (", texture, ")")) |>
  ggplot(aes(h + 0.1, theta)) +
  geom_point(size = 2) +
  scale_x_log10(labels = scales::label_log()) +
  facet_wrap(~label) +
  labs(
    x = "Pressure head h (cm, log scale)",
    y = expression(theta~(m^3/m^3)),
    title = "Synthetic retention observations — two example pedons"
  )
```

---

## 3  Fit Van Genuchten parameters per pedon

### 3.1  Sequential baseline

```{r fit-seq}
t_seq <- system.time(
  fits_seq <- suppressWarnings(
    obs_data |>
      group_by(pedon_id, x, y, texture) |>
      fit_swrc(theta_col = theta, h_col = h, workers = 1L)
  )
)["elapsed"]

cat(sprintf("Sequential: %.2f s for %d pedons (%.1f ms / pedon)\n",
            t_seq, nrow(fits_seq), t_seq / nrow(fits_seq) * 1000))
cat(sprintf("Converged: %d / %d (%.0f %%)\n",
            sum(fits_seq$convergence),
            nrow(fits_seq),
            mean(fits_seq$convergence) * 100))
```

### 3.2  Parallel fit with all physical cores

```{r fit-par}
# R CMD check restricts parallel forking; cap to 2 in that environment
in_check <- isTRUE(as.logical(Sys.getenv("_R_CHECK_LIMIT_CORES_", "FALSE")))
n_cores  <- if (in_check) 2L else parallel::detectCores(logical = FALSE)
cat(sprintf("Workers used: %d\n", n_cores))

t_par <- system.time(
  fits_par <- suppressWarnings(
    obs_data |>
      group_by(pedon_id, x, y, texture) |>
      fit_swrc(theta_col = theta, h_col = h, workers = n_cores)
  )
)["elapsed"]

cat(sprintf("Parallel (%d workers): %.2f s — speedup %.1fx\n",
            n_cores, t_par, t_seq / t_par))
```

### 3.3  Verify consistency

The sequential and parallel paths run identical NLS fits; parameter estimates
must match to machine precision.

```{r verify-consistency}
fits_seq_ord <- arrange(fits_seq, pedon_id)
fits_par_ord <- arrange(fits_par, pedon_id)

max_diff_theta_s <- max(abs(fits_seq_ord$theta_s - fits_par_ord$theta_s),
                        na.rm = TRUE)
max_diff_alpha   <- max(abs(fits_seq_ord$alpha   - fits_par_ord$alpha),
                        na.rm = TRUE)

cat(sprintf("Max |Δθ_s| seq vs par: %.2e\n", max_diff_theta_s))
cat(sprintf("Max |Δα|   seq vs par: %.2e\n", max_diff_alpha))
stopifnot(max_diff_theta_s < 1e-8, max_diff_alpha < 1e-8)
```

---

## 4  Map fitted parameters

With 200 pedons on a regular grid, we can visualise how the fitted θ_s and
shape parameter *n* vary across the landscape.

```{r map-theta-s, fig.alt="Spatial map of fitted saturated water content"}
fits_seq |>
  filter(convergence) |>
  ggplot(aes(x, y, fill = theta_s)) +
  geom_tile(width = 0.95, height = 0.95) +
  scale_fill_distiller(palette = "Blues", direction = 1,
                       name = expression(theta[s]~(m^3/m^3))) +
  coord_equal() +
  labs(
    title = "Fitted saturated water content (θ_s) across the study area",
    x = "Easting (arbitrary)", y = "Northing (arbitrary)"
  ) +
  theme_minimal()
```

```{r map-n, fig.alt="Spatial map of fitted Van Genuchten shape parameter n"}
fits_seq |>
  filter(convergence) |>
  ggplot(aes(x, y, fill = n)) +
  geom_tile(width = 0.95, height = 0.95) +
  scale_fill_distiller(palette = "RdYlBu", direction = 1,
                       name = "n (−)") +
  coord_equal() +
  labs(
    title = "Fitted Van Genuchten shape parameter n",
    subtitle = "Sandy soils (left) have higher n than clay soils (centre-right)",
    x = "Easting (arbitrary)", y = "Northing (arbitrary)"
  ) +
  theme_minimal()
```

---

## 5  Spatial integration with sf (optional)

When pedon coordinates are in a real CRS, the fitted parameters can be
attached directly to an `sf` point layer for further geostatistical analysis
or export as GeoPackage / Shapefile.

```{r sf-section, eval = have_sf}
pedons_sf <- fits_seq |>
  filter(convergence) |>
  sf::st_as_sf(coords = c("x", "y"), crs = NA_integer_)

# Inspect: each feature carries fitted VG parameters
print(pedons_sf[1:4, c("texture", "theta_r", "theta_s", "alpha", "n")])
```

```{r sf-note, eval = !have_sf, echo = FALSE}
message("Install the 'sf' package to run the spatial integration section.")
```

With a real CRS you can then write:

```r
sf::st_write(pedons_sf, "fitted_vg_params.gpkg")
```

---

## 6  Real-world data sources

In practice, the `obs_data` tibble would be built from one of these sources:

| Source | Format | Typical h range | θ unit |
|---|---|---|---|
| **Hypres / UNSODA** | CSV | 0–15 000 cm | m³/m³ |
| **SoilGrids point extraction** | JSON / raster | — | volumetric fraction |
| **Field tensiometer loggers** | CSV / NetCDF | 0–800 hPa | m³/m³ or % |
| **Laboratory pressure-plate** | CSV | 0–15 000 cm | m³/m³ |

Convert pressure in hPa to cm water head with the approximation
1 hPa ≈ 1.02 cm H₂O.

---

## Summary

```{r summary-table, echo = FALSE}
tibble::tribble(
  ~Step, ~Code,
  "Generate observations",     "swrc_van_genuchten() per pedon",
  "Sequential fit (baseline)", "group_by(pedon_id) |> fit_swrc(workers = 1)",
  "Parallel fit",              "fit_swrc(workers = parallel::detectCores(logical = FALSE))",
  "Attach to spatial layer",   "sf::st_as_sf(fits, coords = c('x','y'), crs = ...)",
  "Export",                    "sf::st_write(pedons_sf, 'fitted_vg.gpkg')"
) |>
  knitr::kable(caption = "Multi-pedon workflow steps")
```
