---
title: "Pedotransfer functions: estimating VG parameters from soil texture"
output: rmarkdown::html_vignette
pkgdown:
  downlit: false
vignette: >
  %\VignetteIndexEntry{Pedotransfer functions: estimating VG parameters from soil texture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
```

```{r packages, message=FALSE}
library(tidysoilwater)
library(dplyr)
library(tibble)
library(ggplot2)
```

## What are pedotransfer functions?

Pedotransfer functions (PTFs) are empirical regression equations that
estimate soil hydraulic parameters from readily available soil properties:
sand/silt/clay fractions, organic matter content, and bulk density. They
allow hydraulic characterisation of soils where no retention measurements
exist — the common case in most landscape-scale studies.

This vignette implements the **Wösten et al. (1999)** continuous PTFs for
the Van Genuchten parameter set (θ_r, θ_s, α, n, K_sat) using European
topsoil data.

> Wösten, J. H. M., Lilly, A., Nemes, A., & Le Bas, C. (1999). Development
> and use of a database of hydraulic properties of European soils.
> *Geoderma*, 90(3–4), 169–185.
> <https://doi.org/10.1016/S0016-7061(98)00132-3>

---

## 1. The Wösten (1999) PTF equations

These PTFs use sand (%), silt (%), clay (%), organic matter (% OM), and
bulk density (g/cm³, BD) as predictors.

```{r wosten-ptf}
#' Apply Wösten (1999) continuous PTFs to estimate Van Genuchten parameters.
#'
#' @param sand    Sand content (%)
#' @param silt    Silt content (%)
#' @param clay    Clay content (%)
#' @param om      Organic matter content (%). If unknown, use ~1.0 for
#'                mineral topsoils and ~0.5 for subsoils.
#' @param bd      Bulk density (g/cm³). Typical range 1.0–1.6.
#' @param topsoil Logical; TRUE for A-horizon, FALSE for B/C horizons.
#' @return Tibble with one row of estimated VG parameters.
wosten_ptf <- function(sand, silt, clay, om, bd, topsoil = TRUE) {
  # Notation follows Wösten (1999) Table 4 (continuous PTF)
  ts <- as.integer(topsoil)

  # Saturated water content (cm³/cm³)
  theta_s <- 0.7919 + 0.001691 * clay - 0.29619 * bd -
             0.000001491 * silt^2 + 0.0000821 * om^2 +
             0.02427 / clay + 0.01113 / silt + 0.01472 * log(silt) -
             0.0000733 * om * clay - 0.000619 * bd * clay -
             0.001183 * bd * om - 0.0001664 * topsoil * silt

  # Log alpha (alpha in 1/cm)
  log_alpha <- -14.96 + 0.03135 * clay + 0.0351 * silt +
               0.646 * om + 15.29 * bd - 0.192 * ts -
               4.671 * bd^2 - 0.000781 * clay^2 - 0.00957 * om^2 +
               0.00000235 * silt^2 - 0.221 * om / clay +
               0.02264 * bd * clay + 0.00036 * silt * om -
               0.0791 * bd / om - 0.04991 * silt * bd +
               0.0009936 * silt * om - 0.0002698 * sand * clay -
               0.07215 * log(om)
  alpha <- exp(log_alpha)

  # Log n
  log_n <- -25.23 - 0.02195 * clay + 0.0074 * silt -
           0.1794 * om + 45.5 * bd - 7.24 * bd^2 +
           0.0003658 * clay^2 + 0.002885 * om^2 -
           12.81 / bd - 0.1524 / silt - 0.01958 / om -
           0.2876 * log(silt) - 0.0709 * log(om) -
           44.6 * log(bd) - 0.02264 * bd * clay +
           0.00896 * silt * om + 0.00718 * sand * om
  n <- exp(log_n)

  # Log K_sat (cm/day)
  log_ks <- 7.341 + 0.092 * clay - 0.0031 * silt^2 +
            0.0003881 * clay^2 - 0.3899 * bd -
            0.00251 * silt * clay + 0.0003048 * silt^2 -
            0.001 * clay^2 + 0.0023 * silt - 0.04673 * om +
            0.01559 * ts + 0.00959 * log(silt) - 0.08741 * log(om)
  ks <- exp(log_ks)

  tibble::tibble(
    theta_r = 0,          # Wösten (1999) sets theta_r = 0 for continuous PTF
    theta_s = theta_s,
    alpha   = alpha,
    n       = n,
    ks      = ks
  )
}
```

> **Note:** The Wösten (1999) continuous PTF sets θ_r = 0, which is
> appropriate for the equation set but differs from the full Van Genuchten
> formulation. Use it as-is for comparative purposes; for fitting, a small
> positive θ_r can be assumed (e.g. 0.02–0.05 for mineral soils).

---

## 2. Apply PTF to a soil survey dataset

```{r survey-data}
# Representative soil survey data for five profiles
survey <- tribble(
  ~site,             ~sand, ~silt, ~clay, ~om,  ~bd,
  "Sandy ridge",       78,    12,    10,  1.5,  1.55,
  "Sandy loam slope",  62,    22,    16,  2.1,  1.48,
  "Loam terrace",      42,    38,    20,  2.8,  1.40,
  "Clay loam flat",    28,    38,    34,  3.2,  1.30,
  "Heavy clay valley", 12,    28,    60,  4.5,  1.18
)

# Apply PTF row by row (small dataset — purrr approach)
ptf_params <- purrr::pmap_dfr(
  survey,
  function(site, sand, silt, clay, om, bd) {
    ptf <- wosten_ptf(sand, silt, clay, om, bd, topsoil = TRUE)
    bind_cols(tibble(site = site), ptf)
  }
) |>
  # The VG model requires n > 1. The Wösten (1999) PTF can predict n ≤ 1 for
  # heavy clay soils; clamp to 1.001 as a practical lower bound and flag.
  mutate(
    n_raw = n,
    n     = pmax(n, 1.001),
    ptf_n_clamped = n_raw < 1.001
  )

# Report any clamped rows
if (any(ptf_params$ptf_n_clamped)) {
  message(sprintf(
    "Note: n clamped to 1.001 for %d site(s) where PTF predicted n ≤ 1:\n  %s",
    sum(ptf_params$ptf_n_clamped),
    paste(ptf_params$site[ptf_params$ptf_n_clamped], collapse = ", ")
  ))
}

ptf_params <- ptf_params |> select(-n_raw, -ptf_n_clamped)
ptf_params
```

---

## 3. Generate SWRC curves from PTF parameters

```{r swrc-from-ptf}
heads <- 10^seq(0, 5, by = 0.1)

ptf_curves <- ptf_params |>
  tidyr::crossing(h = heads) |>
  mutate(m = 1 - 1 / n) |>
  swrc_van_genuchten(
    theta_r = theta_r,
    theta_s = theta_s,
    alpha   = alpha,
    n       = n,
    h       = h
  )
```

```{r ptf-swrc-plot}
ggplot(ptf_curves, aes(x = h, y = .theta, colour = site)) +
  geom_line(linewidth = 0.9) +
  scale_x_log10(labels = scales::label_comma()) +
  labs(
    x      = "Matric potential |h| (cm)",
    y      = expression(theta~(m^3/m^3)),
    colour = "Site",
    title  = "PTF-estimated soil water retention curves",
    caption = "Wösten (1999) continuous PTF"
  )
```

---

## 4. Derive PAWC from PTF parameters

```{r pawc-ptf}
pawc_ptf <- ptf_params |>
  tidyr::crossing(h = c(330, 15000)) |>
  mutate(m = 1 - 1 / n) |>
  swrc_van_genuchten(theta_r = theta_r, theta_s = theta_s,
                     alpha = alpha, n = n, h = h) |>
  mutate(label = if_else(h == 330, "theta_fc", "theta_wp")) |>
  select(site, label, .theta) |>
  tidyr::pivot_wider(names_from = label, values_from = .theta) |>
  mutate(pawc = theta_fc - theta_wp)

pawc_ptf |>
  select(site, theta_fc, theta_wp, pawc) |>
  arrange(desc(pawc))
```

```{r pawc-bar}
ggplot(pawc_ptf,
       aes(x = reorder(site, pawc), y = pawc)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    x = NULL,
    y = "PAWC (m³/m³)",
    title = "Plant-available water capacity from PTF estimates"
  )
```

---

## 5. PTF vs fitted: how well do they compare?

Where measured (h, θ) data exist, it is worth comparing PTF estimates to
fitted parameters.

```{r ptf-vs-fitted}
set.seed(3109)

# Simulate lab-measured retention for the loam terrace (site 3)
loam_true <- ptf_params |> filter(site == "Loam terrace")
lab_heads  <- c(0, 10, 33, 100, 330, 1000, 5000, 15000)
lab_obs <- tibble(
  h     = lab_heads,
  theta = with(loam_true,
    theta_r + (theta_s - theta_r) /
      (1 + (alpha * lab_heads)^n)^(1 - 1/n)
  ) + rnorm(length(lab_heads), 0, 0.008)
)

# Fit VG to the simulated lab data
fitted <- fit_swrc(lab_obs, theta_col = theta, h_col = h)

comparison <- bind_rows(
  mutate(loam_true |> select(theta_r, theta_s, alpha, n), source = "PTF"),
  mutate(fitted     |> select(theta_r, theta_s, alpha, n), source = "Fitted")
) |>
  filter(!is.na(theta_r), theta_s > theta_r, n > 1)

comparison
```

```{r ptf-vs-fit-plot}
curves_comparison <- comparison |>
  tidyr::crossing(h = heads) |>
  mutate(m = 1 - 1 / n) |>
  swrc_van_genuchten(theta_r = theta_r, theta_s = theta_s,
                     alpha = alpha, n = n, h = h)

ggplot() +
  geom_line(data  = curves_comparison,
            aes(x = h, y = .theta, colour = source, linetype = source),
            linewidth = 0.9) +
  geom_point(data = lab_obs, aes(x = h, y = theta), size = 2.5) +
  scale_x_log10(labels = scales::label_comma()) +
  labs(
    x      = "Matric potential |h| (cm)",
    y      = expression(theta~(m^3/m^3)),
    colour = "Source", linetype = "Source",
    title  = "Loam terrace: PTF estimate vs NLS fit to lab data",
    caption = "Points = simulated lab observations"
  )
```

---

## 6. Real-world PTF data sources

| Source | Parameters | Coverage | Notes |
|--------|-----------|----------|-------|
| **Wösten et al. (1999)** | θ_s, α, n, K_sat | European soils | Used in this vignette |
| **Rawls & Brakensiek (1985)** | θ_r, θ_s, α, n | US soils | Class-mean or continuous |
| **Schaap et al. (2001) — Rosetta** | θ_r, θ_s, α, n, K_sat | Global | Neural network; `rosettaAPI` R package |
| **SoilGrids + Cosby (1984)** | B-C parameters | Global 250m | Via `soilDB` R package |

For most applications, **Rosetta 3** (Schaap 2001, updated Zhang & Schaap
2017) provides the best accuracy for global use. The `rosettaAPI` package
provides access to the web API.

---

## Summary

The PTF workflow integrates seamlessly with `tidysoilwater`:

```r
soil_data |>
  mutate(wosten_ptf(sand, silt, clay, om, bd)) |>   # add PTF columns
  crossing(h = heads) |>
  mutate(m = 1 - 1 / n) |>
  swrc_van_genuchten(theta_r, theta_s, alpha, n, h)  # evaluate curve
```

PTF-derived parameters are best treated as priors — useful for rapid
assessment but best validated against local laboratory data when available.
