---
title: "Soil water retention with swrc_van_genuchten()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Soil water retention with swrc_van_genuchten()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
```

```{r packages, message=FALSE}
library(tidysoilwater)
library(dplyr)
library(tibble)
library(ggplot2)
```

## Overview

`swrc_van_genuchten()` computes the volumetric water content θ at a given
matric potential h using the Van Genuchten (1980) closed-form model with the
Mualem (1976) constraint (m = 1 − 1/n):

$$\theta(h) = \theta_r + \frac{\theta_s - \theta_r}{\left(1 + (\alpha |h|)^n\right)^{1 - 1/n}}$$

---

## 1. Parameters

| Parameter | Symbol | Units | Typical range | Meaning |
|-----------|--------|-------|---------------|---------|
| `theta_r` | θ_r | m³/m³ | 0.01 – 0.10 | Residual water content |
| `theta_s` | θ_s | m³/m³ | 0.30 – 0.65 | Saturated water content |
| `alpha`   | α   | 1/cm  | 0.005 – 0.15 | Inverse of air-entry pressure |
| `n`       | n   | –     | 1.1 – 3.5    | Pore-size distribution index |
| `h`       | h   | cm    | 0 – 150 000  | Matric potential (absolute value used) |

The derived parameter m = 1 − 1/n is handled internally; you do not need to
supply it.

---

## 2. Single-soil curve

```{r single-soil}
# Loam soil (Carsel & Parrish 1988 typical values)
loam <- tibble(h = c(0, 10, 33, 100, 330, 1000, 3300, 5000, 10000, 15000))

loam_swrc <- swrc_van_genuchten(loam,
  theta_r = 0.078, theta_s = 0.430,
  alpha   = 0.036, n = 1.56, h = h)

loam_swrc
```

```{r single-plot}
ggplot(loam_swrc, aes(x = h, y = .theta)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_log10(labels = scales::label_comma()) +
  labs(
    x     = "Matric potential |h| (cm)",
    y     = expression(theta~(m^3/m^3)),
    title = "Soil water retention curve — loam"
  )
```

Key reference thresholds on this curve:

```{r thresholds}
swrc_van_genuchten(
  tibble(h = c(0, 33, 330, 15000), label = c("saturation", "field capacity",
                                               "intermediate", "wilting point")),
  theta_r = 0.078, theta_s = 0.430,
  alpha = 0.036, n = 1.56, h = h
) |>
  select(label, h, .theta)
```

---

## 3. Comparing texture classes

Standard Carsel & Parrish (1988) parameters for five common texture classes:

```{r texture-params}
textures <- tribble(
  ~texture,        ~theta_r, ~theta_s, ~alpha,    ~n,
  "sand",            0.045,    0.430,   0.145,  2.68,
  "sandy loam",      0.065,    0.410,   0.075,  1.89,
  "loam",            0.078,    0.430,   0.036,  1.56,
  "clay loam",       0.095,    0.410,   0.019,  1.31,
  "clay",            0.100,    0.380,   0.015,  1.09
)
```

```{r texture-curves}
heads <- 10^seq(0, 5, by = 0.05)

curves <- textures |>
  tidyr::crossing(h = heads) |>
  swrc_van_genuchten(theta_r = theta_r, theta_s = theta_s,
                     alpha = alpha, n = n, h = h)

ggplot(curves, aes(x = h, y = .theta, colour = texture)) +
  geom_line(linewidth = 0.9) +
  scale_x_log10(labels = scales::label_comma()) +
  labs(
    x      = "Matric potential |h| (cm)",
    y      = expression(theta~(m^3/m^3)),
    colour = "Texture",
    title  = "Soil water retention curves by texture class"
  )
```

---

## 4. Tidy evaluation: columns or scalars

All parameters accept either a bare column name from `data` or a scalar
numeric value — mix and match as needed.

```{r tidyeval-demo}
# Per-row alpha and n; scalar theta_r and theta_s
param_grid <- tibble(
  alpha = c(0.02, 0.05, 0.10, 0.15),
  n     = c(1.25, 1.50, 2.00, 2.68)
) |>
  tidyr::crossing(h = c(33, 330, 15000))

swrc_van_genuchten(param_grid,
  theta_r = 0.05,     # scalar — same for all rows
  theta_s = 0.42,     # scalar
  alpha   = alpha,    # from data column
  n       = n,        # from data column
  h       = h
) |>
  select(alpha, n, h, .theta)
```

---

## 5. Parameter sensitivity

### Effect of α (air-entry / pore-size)

Higher α → curve drops earlier (coarser pores drain at lower suction).

```{r alpha-sensitivity}
tribble(
  ~alpha, ~label,
  0.005,  "α = 0.005 (fine)",
  0.02,   "α = 0.02",
  0.05,   "α = 0.05",
  0.145,  "α = 0.145 (coarse)"
) |>
  tidyr::crossing(h = heads) |>
  swrc_van_genuchten(theta_r = 0.05, theta_s = 0.42,
                     alpha = alpha, n = 1.8, h = h) |>
  ggplot(aes(x = h, y = .theta, colour = label)) +
  geom_line(linewidth = 0.9) +
  scale_x_log10(labels = scales::label_comma()) +
  labs(x = "|h| (cm)", y = expression(theta), colour = "α value",
       title = "Sensitivity to α (n fixed at 1.8)")
```

### Effect of n (curve shape)

Higher n → steeper, more S-shaped curve (well-sorted pore-size distribution).

```{r n-sensitivity}
tribble(
  ~n,    ~label,
  1.10,  "n = 1.10 (clay-like)",
  1.30,  "n = 1.30",
  1.80,  "n = 1.80",
  2.68,  "n = 2.68 (sand-like)"
) |>
  tidyr::crossing(h = heads) |>
  swrc_van_genuchten(theta_r = 0.05, theta_s = 0.42,
                     alpha = 0.04, n = n, h = h) |>
  ggplot(aes(x = h, y = .theta, colour = label)) +
  geom_line(linewidth = 0.9) +
  scale_x_log10(labels = scales::label_comma()) +
  labs(x = "|h| (cm)", y = expression(theta), colour = "n value",
       title = "Sensitivity to n (α fixed at 0.04)")
```

---

## 6. Sign convention

Both positive and negative h values are accepted — the absolute value is
used internally. This means pF-style (positive suction) and matric potential
(negative pressure) data both work without conversion.

```{r sign-convention}
tibble(h_negative = c(-0, -10, -100, -1000, -15000)) |>
  swrc_van_genuchten(theta_r = 0.05, theta_s = 0.42,
                     alpha = 0.04, n = 1.8, h = h_negative)
```

---

## 7. Performance

The function uses direct tibble assignment with scalar recycling — no loops,
no intermediate allocations.

```{r perf}
df_1m <- tibble(h = rep(c(33, 330, 15000), length.out = 1e6))
elapsed <- system.time(
  swrc_van_genuchten(df_1m, theta_r = 0.05, theta_s = 0.42,
                     alpha = 0.04, n = 1.8, h = h)
)["elapsed"]
cat(sprintf("1 M rows in %.0f ms\n", elapsed * 1000))
```

---

## See also

* `vignette("hydraulic-conductivity")` — compute K(h) from the same parameters
* `vignette("fitting-diagnostics")` — fit parameters from observed data
* `vignette("multi-pedon-workflow")` — grouped and parallel fitting at scale
